Virtualization of a Real-Time Operating System for Robot Control with a Focus on Real-Time Compliance

### 2024-08-30 Masterarbeit_Robotik_Pamuk.tex
bind paper
Paper: Abstract, Introduction, Methodology, Implementation, Results, Discussion, Summary and Outlook, Bibliography

%
% FH Technikum Wien
% !TEX encoding = UTF-8 Unicode
%
% Erstellung von Master- und Bachelorarbeiten an der FH Technikum Wien mit Hilfe von LaTeX und der Klasse TWBOOK
%
% Um ein eigenes Dokument zu erstellen, müssen Sie folgendes ergänzen:
% 1) Mit \documentclass[..] einstellen: Master- oder Bachelorarbeit, Studiengang und Sprache
% 2) Mit \newcommand{\FHTWCitationType}.. Zitierstandard festlegen (wird in der Regel vom Studiengang vorgegeben - bitte erfragen)
% 3) Deckblatt, Kurzfassung, etc. ausfüllen
% 4) und die Arbeit schreiben (die verwendeten Literaturquellen in Literature.bib eintragen)
%
% Getestet mit TeXstudio mit Zeichenkodierung ISO-8859-1 (=ansinew/latin1) und MikTex unter Windows
% Zu beachten ist, dass die Kodierung der Datei mit der Kodierung des paketes inputenc zusammen passt!
% Die Kodierung der Datei twbook.cls MUSS ANSI betragen!
% Bei der Verwendung von UTF8 muss dnicht nur die Kodierung des Dokuments auf UTF8 gestellt sein, sondern auch die des BibTex-Files!
%
% Bugreports und Feedback bitte per E-Mail an latex@technikum-wien.at
%
% Versionen
% *) V0.7: 9.1.2015, RO: Modeline angepasst und verschoben
% *) V0.6: 10.10.2014, RO: Weitere Anpassung an die UK
% *) V0.5: 8.8.2014, WK: Literaturquellen überarbeitet und angepasst
% *) V0.4: 4.8.2014, WK: Initalversion in SVN eingespielt
%
\documentclass[MMR,Master,english]{twbook}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
%
% Hier biblatex & Biber konfigurieren; Vergessen Sie nicht, dass Sie biber verwenden müssen um eine Bibliothek zu erzeugen
%
\usepackage[backend=biber, style=numeric, sorting=none]{biblatex}
\addbibresource{Literature.bib}
%
% Bei Bedarf bitte hier die Syntax-Highlightings anpassen
%
% own 
\usepackage{hyperref}
\usepackage{float}
\usepackage{comment}
\usepackage{csquotes}
\usepackage{pgfplots}
\usepackage{soul}
\usepackage{color}
\usepackage{xcolor}  % Required for custom colors
\sethlcolor{blue}
\setcounter{secnumdepth}{3} % in text 
\setcounter{tocdepth}{3} % contents
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{rotating}
\usepackage{tikz}
\usepackage{setspace}
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}

@@

\bigskip \noindent Altogether, this thesis provides a comprehensive blueprint for making a virtualized guest system real-time capable in a host system with deterministic behavior.}
\keywords{Virtualization, Real-Time Systems, Latency Reduction, Robotic Control}

@@

\section{Approach}\label{sec:approach}
The goal is to The \texttt{latency} program is first executed in the bare metal Salamander 4 OS.
The goal is to virtualize the Salamander 4 OS and bring its real-time performance closer to that of the hardware version and guarantee deterministic and reliable behavior. For that purpose, the first step is to apply the PREEMPT-RT patch to the host system and subsequently tune the real-time performance through BIOS, kernel, host, guest and QEMU/KVM configurations. These modifications are added sequentially and tested together. First, the BIOS is configured, followed by the kernel. The BIOS settings are not reverted when moving to the kernel configurations. Next, the host is configured, but the BIOS and kernel settings remain unchanged. After that, the guest is configured, and finally, QEMU/KVM settings are applied. This means that the configurations are not isolated but are applied and then tested as a whole. The key metric here is the latency of the system, and the real-time performance is evaluated using the \texttt{latency} tool of the Xenomai suite after each configuration step. At the end, the results are domenstarted alltogether and a holistic comparison and discussion is carried out. In addition, the improvement in real-time performance and determinism is demonstrated using a robotic application, where the modified virtualization is compared with the hardware version.

@@

\begin{comment}

\chapter{To Include}\label{cha:to_include}

\section{KVM Entry and KVM Exit}\label{subsubsec:kvm_entry_exit}
\textcolor{red}{\noindent Upon isolating a CPU to the QEMU process, it was anticipated that the guest would utilize nearly 100\% of the CPU's capacity, with minimal to no intervention from the host. However, the isolcpus function only isolates CPUs from the scheduler, meaning that user-space processes won't be scheduled on these CPUs, but kernel tasks and interrupts can still utilize the CPU.This led to the investigation of the causes for the observed high and inconsistent latency. The guest operates within the~\texttt{kvm\_entry} and~\texttt{kvm\_exit} events of the host. Kernelshark revealed a high frequency of~\texttt{kvm\_exit} events, indicating that the guest frequently relinquishes control of the CPU back to the host. This frequent switching hinders the guest's ability to run continuously, thereby increasing the virtualization latency.}

\textcolor{red}{To further understand this, trace-cmd was employed to trace various events in the host-guest communication, including the reasons for these events. Specifically, the causes for~\texttt{kvm\_exit} events were analyzed. The command sudo trace-cmd record -e all -A @3:823 --name Salamander4 -e all was executed on the host for a duration of 5 seconds. The results in Figure~\ref{fig:kvm_exit_taskset} were obtained. Additionally, Table~\ref{tab:kvm_exit} provides a short description of the observed~\texttt{kvm\_exit} events.}

\vspace{2em}
\begin{table}[H]
	\centering
	\begin{tabular}{|l|p{0.62\textwidth}|}
		\hline
		\textbf{Exit Reason} & \textbf{Description}                                     \\ \hline
		APIC\_WRITE          & Triggered when the guest writes to its APIC.             \\ \hline
		EXTERNAL\_INTERRUPT  & Triggered by external hardware interrupts.               \\ \hline
		HLT                  & Triggered when the guest executes the HLT instruction.   \\ \hline
		EPT\_MISCONFIG       & Triggered by a misconfiguration in the EPT.              \\ \hline
		PREEMPTION\_TIMER    & Triggered when the host's preemption timer expires.      \\ \hline
		PAUSE\_INSTRUCTION   & Triggered when the PAUSE instruction is executed.        \\ \hline
		EPT\_VIOLATION       & Triggered by a violation of the EPT permission settings. \\ \hline
		IO\_INSTRUCTION      & Triggered when the guest executes an I/O instruction.    \\ \hline
		EOI\_INDUCED         & Triggered when an EOI signal is sent to the APIC.        \\ \hline
		MSR\_READ            & Triggered when the guest reads from a MSR.               \\ \hline
		CPUID                & Triggered when the guest executes the CPUID instruction. \\ \hline
	\end{tabular}
	\caption{Description of kvm\_exit reasons}
	\label{tab:kvm_exit}
\end{table}

\textcolor{red}{\noindent In the process of analyzing the~\texttt{kvm\_exit} events, several reasons for these exits were identified. The most frequent among these were the~\texttt{APIC\_WRITE} and~\texttt{HLT} events. The former is initiated when the guest writes to its Advanced Programmable Interrupt Controller (APIC), a component of the CPU that manages hardware interrupts. The latter occurs when the guest executes the HLT instruction, effectively halting the CPU until the next external interrupt is fired. Other significant but less frequent events included~\texttt{EXTERNAL\_INTERRUPT} and~\texttt{IO\_INSTRUCTION}. These events are indicative of the guest's interaction with hardware devices and its execution of I/O operations. Events such as~\texttt{EPT\_MISCONFIG} and~\texttt{PREEMPTION\_TIMER} were also noted. These could potentially signal issues with memory management and the host's scheduling of the guest. While events like~\texttt{PAUSE\_INSTRUCTION},~\texttt{EPT\_VIOLATION},~\texttt{EOI\_INDUCED},~\texttt{MSR\_READ}, and~\texttt{CPUID} were the least frequent, they still provide valuable insights into the guest's behavior and the host-guest interaction.}

\textcolor{red}{\noindent When the script is started from the host, the QEMU process can be scheduled to run on any available core, as it is not bound to a specific CPU core. This means that the QEMU process may frequently switch between different cores, leading to an increase in latency. As the goal was to reduce latency in the guest, the first step was to isolate a CPU of the host and dedicate it solely to the QEMU process, so that it cannot be used for other tasks on the user level. Without CPU isolation, context switches take place at the operating system level and not at the hypervisor level. This explains why there are fewer~\texttt{kvm\_exit} events. However, this will also, as previously shown, lead to higher latency, as context switches at the operating system level generally take longer than a~\texttt{kvm\_exit} and ~\texttt{kvm\_entry}. When the CPU was dedicated to the QEMU process, on the other hand, there was a significant increase in~\texttt{kvm\_exit} events. This is because every context switch takes place at the hypervisor level. Nevertheless, lower latency was achieved thereby, as the qemu process is no longer influenced by the CPU scheduling of the operating system.}

``korrekt''

\noindent \enquote{This LaTeX.}

\section{Trace-cmd and Kernelshark}
After analyzing the inital latency of both versions, Trace-cmd and Kernelshark were used to further inspect the reasons that caused this divergence.

\section{Host and Guest tasks}
\vspace{2em}
\begin{table}[H]
\centering
\begin{minipage}{.5\textwidth}
\centering
\caption{Host report CPU19 (Total of 445.908)}
\label{tab:results_host_report}
\begin{tabular}{|c|c|c|}
	\hline
	PID    & Task                          & Count   \\ \hline
	182579 & qemu-system-x86               & 302.748 \\ \hline
	0      & \textless{}idle\textgreater{} & 112.911 \\ \hline
	182618 & vhost                         & 21.204  \\ \hline
	182572 & qemu-system-x86               & 7.597   \\ \hline
	182755 & qemu-system-x86               & 644     \\ \hline
	182754 & qemu-system-x86               & 643     \\ \hline
	181870 & kworker/19:1                  & 139     \\ \hline
	3820   & kworker/19:1H                 & 16      \\ \hline
	94     & migration/19                  & 6       \\ \hline
\end{tabular}
%
\begin{minipage}{.5\textwidth}
\centering
\caption{Guest report (Total of 362.370)}
\label{tab:results_guest_report}
\begin{tabular}{|c|c|c|}
	\hline
	PID & Task                          & Count   \\ \hline
	0   & \textless{}idle\textgreater{} & 150.744 \\ \hline
	331 & LRT-Main                      & 56.697  \\ \hline
	377 & trace-cmd                     & 48.507  \\ \hline
	346 & CLI                           & 26.426  \\ \hline
	378 & kthreadd                      & 25.837  \\ \hline
	340 & MainTaskLow                   & 19.291  \\ \hline
	339 & \textless{}...\textgreater{}  & 9.980   \\ \hline
	34  & MainTaskHigh                  & 9.185   \\ \hline
	327 & LE-Logger                     & 4.965   \\ \hline
	369 & kworker/0:0                   & 2.793   \\ \hline
	321 & kWorker-LRT                   & 2.542   \\ \hline
	328 & LRTMgr-Main                   & 1.651   \\ \hline
	34  & LrtMgrCyclic                  & 1.220   \\ \hline
	332 & cobalt\_printf                & 1.112   \\ \hline
	325 & LE-System                     & 534     \\ \hline
	343 & TCP-Listen                    & 187     \\ \hline
	15  & rcu\_preempt                  & 162     \\ \hline
	25  & kcompactd0                    & 122     \\ \hline
	58  & kworker/0:1H                  & 96      \\ \hline
	63  & kworker/u2:2                  & 89      \\ \hline
	8   & jbd2/sda-8                    & 86      \\ \hline
	22  & kworker/0:1                   & 56      \\ \hline
	1   & init                          & 31      \\ \hline
	2   & kthreadd                      & 25      \\ \hline
	375 & trace-cmd                     & 24      \\ \hline
	14  & ksoftirqd/0                   & 8       \\ \hline
\end{tabular}

\end{table}

\noindent In the following, the host and guest tasks along with their impact on system latency are briefly described.

\begin{itemize}
	\item \textbf{qemu-system-x86}: Part of the QEMU process and specifically, this task emulates x86 systems. In Table~\ref{tab:results_host_report}, it occurs four times under different PIDs, hence there are four threads of it.
	\item \textbf{<idle>}: This represents the idle time of the CPU, hence it is not being used by any process, allowing to save power. The system halts until the next interrupt, which could be a timer interrupt, I/O interrupt, etc.
	\item \textbf{vhost}: A kernel module which improves virtual input/output (virtio) performance by handling virtqueues in the kernel, thereby reducing context switches and system calls.
	\item \textbf{kworker/19:1}: A kernel worker thread created by the Linux kernel, kworker/19:1 performs work in response to system events. The number after the slash and colon indicate the CPU core and internal ID of the worker thread, respectively.
	\item \textbf{kworker/19:1H}: Similar to kworker/19:1, kworker/19:1H is a kernel worker thread, with the ‘H’ suggesting that this thread handles hardware interrupts.
	\item \textbf{migration/19}: The migration process is a kernel process that balances load across CPU cores by moving threads from one CPU to another. The number after the slash indicates the CPU core to which the migration process is bound. (URL: \url{https://elixir.bootlin.com/linux/latest/source/kernel/sched/core.c#L2325})

The Figure below (deleted) compares non-optimized guest latency with optimized guest latency and includes optimized bare metal latency as a reference. The data shows that a 40\% reduction in QD1 latency is achievable through system tuning.

\end{comment}

@@

\noindent The improved latency was also tested with the robotic application in Subsection~\ref{sec:robotic_application}. The difference between the command time and the time the signal reaches the PWM was measured 1,000 times for the untuned, tuned and hardware version of Salamander 4. The results are visualized in Figure~\ref{fig:combined_latencies}.

@@

\begin{table}[H]
	\small
	\centering
	\caption{Comparison of Robotic Application Latency Statistics}
	\label{tab:robotic_application_latency_values_combined}
	\setlength{\tabcolsep}{0.5em} % for the horizontal padding
	{\renewcommand{\arraystretch}{1.2}% for the vertical padding
		\begin{tabular}{|l|l|l|l|l|l|}
			\hline
			\textbf{Tuning} & \textbf{Samples} & \textbf{Lat Min (ms)} & \textbf{Lat Avg (ms)} & \textbf{Lat Max (ms)} & \textbf{Std Dev (ms)} \\ \hline
			Hardware & 1000 & 1.211 & 1.347 & 1.49 & 0.082 \\ \hline
			Untuned Virtualization & 1000 & 3.1 & 24.603 & 129.46 & 13.876 \\ \hline
			Tuned Virtualization & 1000 & 1.219 & 2.62 & 3.988 & 0.812 \\ \hline
		\end{tabular}}
\end{table}

@@

\chapter{Summary and Outlook}\label{cha:summary_and_outlook}

Hardware-based systems are limited in adaptability and can be costly, especially when scaling operations. Physical access for updates and maintenance is challenging, leading to downtime and lost productivity. While virtualization addresses these issues, it introduces increased overhead and latency.

>>>>

%
% FH Technikum Wien
% !TEX encoding = UTF-8 Unicode
%
% Erstellung von Master- und Bachelorarbeiten an der FH Technikum Wien mit Hilfe von LaTeX und der Klasse TWBOOK
%
% Um ein eigenes Dokument zu erstellen, müssen Sie folgendes ergänzen:
% 1) Mit \documentclass[..] einstellen: Master- oder Bachelorarbeit, Studiengang und Sprache
% 2) Mit \newcommand{\FHTWCitationType}.. Zitierstandard festlegen (wird in der Regel vom Studiengang vorgegeben - bitte erfragen)
% 3) Deckblatt, Kurzfassung, etc. ausfüllen
% 4) und die Arbeit schreiben (die verwendeten Literaturquellen in Literature.bib eintragen)
%
% Getestet mit TeXstudio mit Zeichenkodierung ISO-8859-1 (=ansinew/latin1) und MikTex unter Windows
% Zu beachten ist, dass die Kodierung der Datei mit der Kodierung des paketes inputenc zusammen passt!
% Die Kodierung der Datei twbook.cls MUSS ANSI betragen!
% Bei der Verwendung von UTF8 muss dnicht nur die Kodierung des Dokuments auf UTF8 gestellt sein, sondern auch die des BibTex-Files!
%
% Bugreports und Feedback bitte per E-Mail an latex@technikum-wien.at
%
% Versionen
% *) V0.7: 9.1.2015, RO: Modeline angepasst und verschoben
% *) V0.6: 10.10.2014, RO: Weitere Anpassung an die UK
% *) V0.5: 8.8.2014, WK: Literaturquellen überarbeitet und angepasst
% *) V0.4: 4.8.2014, WK: Initalversion in SVN eingespielt
%
\documentclass[MMR,Master,english]{style/twbook}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
%
% Hier biblatex & Biber konfigurieren; Vergessen Sie nicht, dass Sie biber verwenden müssen um eine Bibliothek zu erzeugen
%
\usepackage[backend=biber, style=numeric, sorting=none]{biblatex}
\addbibresource{literature/Literature.bib}
%
% Bei Bedarf bitte hier die Syntax-Highlightings anpassen
%
% own 
\usepackage{hyperref}
\usepackage{float}
\usepackage{comment}
\usepackage{csquotes}
\usepackage{pgfplots}
\usepackage{soul}
\usepackage{color}
\usepackage{xcolor}  % Required for custom colors
\sethlcolor{blue}
\setcounter{secnumdepth}{3} % in text 
\setcounter{tocdepth}{3} % contents
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{rotating}
\usepackage{tikz}
\usepackage{setspace}
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}

@@

\bigskip \noindent Altogether, this thesis provides a comprehensive blueprint for making a virtualized guest system real-time capable in a host system with deterministic behavior.}
\keywords{Virtualization, Real-Time Systems, Latency Reduction, Robot Control}

@@

\section{Approach}\label{sec:approach}
The goal is to The \texttt{latency} program is first executed in the bare metal Salamander 4 OS.
The goal is to virtualize the Salamander 4 OS and bring its real-time performance closer to that of the hardware version and guarantee deterministic and reliable behavior. For that purpose, the first step is to apply the PREEMPT-RT patch to the host system and subsequently tune the real-time performance through BIOS, kernel, host, guest and QEMU/KVM configurations. These modifications are added sequentially and tested together. First, the BIOS is configured, followed by the kernel. The BIOS settings are not reverted when moving to the kernel configurations. Next, the host is configured, but the BIOS and kernel settings remain unchanged. The guest OS is already based on Xenomai and therefore, no additional modifications are necessary for the guest OS itself. Finally, QEMU/KVM settings are applied. This means that the configurations are not isolated but are applied and then tested as a whole. The key metric here is the latency of the system, and the real-time performance is evaluated using the \texttt{latency} tool of the Xenomai suite after each configuration step. At the end, the results are domenstarted alltogether and a holistic comparison and discussion is carried out. In addition, the improvement in real-time performance and determinism is demonstrated using a robotic application, where the modified virtualization is compared with the hardware version.

@@

@@

\noindent The improved latency was also tested with the robotic application in Subsection~\ref{sec:robotic_application}. The difference between the command time and the time the signal reaches the PWM was measured 1,000 times for the bare metal, untuned and tuned version of Salamander 4. The results are visualized in Figure~\ref{fig:combined_latencies}.

@@

\begin{table}[H]
	\small
	\centering
	\caption{Comparison of Robotic Application Latency Statistics}
	\label{tab:robotic_application_latency_values_combined}
	\setlength{\tabcolsep}{0.5em} % for the horizontal padding
	{\renewcommand{\arraystretch}{1.2}% for the vertical padding
		\begin{tabular}{|l|l|l|l|l|l|}
			\hline
			\textbf{Tuning} & \textbf{Samples} & \textbf{Lat Min (ms)} & \textbf{Lat Avg (ms)} & \textbf{Lat Max (ms)} & \textbf{Std Dev (ms)} \\ \hline
			Bare Metal & 1000 & 1.211 & 1.347 & 1.49 & 0.082 \\ \hline
			Untuned Virtualization & 1000 & 3.1 & 24.603 & 129.46 & 13.876 \\ \hline
			Tuned Virtualization & 1000 & 1.219 & 2.62 & 3.988 & 0.812 \\ \hline
		\end{tabular}}
\end{table}

@@

\chapter{Summary and Outlook}\label{cha:summary_and_outlook}
Hardware-based systems are limited in adaptability and can be costly, especially when scaling operations. Physical access for updates and maintenance is challenging, leading to downtime and lost productivity. While virtualization addresses these issues, it introduces increased overhead and latency.

### 2024-08-30 Paper.tex
Paper: Abstract, Introduction, Methodology, Implementation, Results, Discussion, Summary and Outlook, Bibliography

% Zitierung im Harvard-Style
\usepackage{harvard}
\usepackage{url} %Darstellung von URLs erlauben
\citationstyle{dcu}  % für Zitate
\bibliographystyle{style/HarvardFHTWMR_V1_2e}% für Literaturverzeichnis
\citationmode{abbr}
\newcommand\citet{\citeasnoun}
\newcommand\citep{\cite}
% \citet(cootes01) -> Cootes (2001)
% \citep(cootes01) -> (Cootes, 2001)
\renewcommand{\harvardand}{\&}
%Deutsch (wahlweise mit DE)
\newcommand{\acessedthrough}{Verfügbar unter:}%Für URL-Angabe
\newcommand{\acessedthroughp}{Verfügbar durch:}%Für URL-Angabe (Geschützte Datenbank, Zugriff durch FH)
\newcommand{\acessedat}{Zugang am}%Für URL-Datum-Angabe
%Englisch (wahlweise mit EN)
%\newcommand{\acessedthrough}{Available at:}%Für URL-Angabe
%\newcommand{\acessedthroughp}{Available through:}%Für URL-Angabe (Geschützte Datenbank, Zugriff durch FH)
%\newcommand{\acessedat}{Accessed}%Für URL-Datum-Angabe
% bis hierher

@@

@@

  \begin{multicols}{2}
    \abstract{\textbf{Virtualizing real-time operating systems for robotic control offers many advantages over traditional hardware-based approaches, especially when it comes to flexibility, scaling, and costs. Virtualization addresses these issues but introduces overhead and latency. This thesis investigates the virtualization of a real-time operating system, which is built with Yocto, employs hard real-time through Xenomai 3 and is emulated using QEMU/KVM. The primary objective aim was to bridge the latency gap between virtualized and bare metal versions to ensure deterministic and reliable behavior, which is crucial for real-time robotics. Initial latency measurements showed a high gap between bare metal and virtualized setups. Extensive tuning across BIOS, kernel, host OS, QEMU/KVM, and Salamander 4 (guest) reduced worst-case latency from 707.622 µs to 17.134 µs, close to the bare metal performance of 10.709 µs. The improvement is validated using a robotic application, comparing tuned virtualization with untuned and hardware versions.}} {\textbf{Virtualization, Real-Time Systems, Latency Reduction, Robot Control}}

    \section{Introduction}
    \begin{figure}[H]
      \centering
      \includegraphics[width=1.0\columnwidth]{img/implementation/gnuplot_combined_max_latency.png}
      \caption[Initial Comparison of Latency Distribution between Hardware and Virtualization]{Initial Comparison of Latency Distribution between Hardware and Virtualization}
      \label{fig:gnuplot_max_latency_combined}
    \end{figure}

    \section{Methodology}
    \section{Implementation}
    \section{Initial Situation} 
    \section{Real-Time Performance Tunings}
    \section{Real-Time Robotic Application}
    \section{Results}
    \begin{figure}[H]
      \centering
      \includegraphics[width=1.0\columnwidth]{img/results/gnuplot_combined_max_latency_all.png}
      \caption[Comparison of Latency Distribution of Salamander 4 Configurations]{Comparison of Latency Distribution of Salamander 4 under different Configurations}
      \label{fig:max_latency_combined_results}
    \end{figure}

    \begin{figure}[H]
      \centering
      \includegraphics[width=1.0\columnwidth]{img/results/combined_latencies.png}
      \caption[Comparison of Robotic Application Latency of Salamander 4 Configurations]{Comparison of Robotic Application Latency of Salamander 4 under different Configurations}
      \label{fig:combined_latencies}
    \end{figure}

\begin{table}[H]
	\scriptsize
	\centering
	\caption{Comparison of Robotic Application Latency Statistics}
	\label{tab:robotic_application_latency_values_combined}
	\setlength{\tabcolsep}{0.5em} % for the horizontal padding
	{\renewcommand{\arraystretch}{1.2}% for the vertical padding
		\begin{tabular}{|l|l|l|l|l|l|}
			\hline
			\textbf{Tuning} & \textbf{Samples} & \textbf{Lat Min (ms)} & \textbf{Lat Avg (ms)} & \textbf{Lat Max (ms)} & \textbf{Std Dev (ms)} \\ \hline
			Hardware & 1000 & 1.211 & 1.347 & 1.49 & 0.082 \\ \hline
			Untuned Virtualization & 1000 & 3.1 & 24.603 & 129.46 & 13.876 \\ \hline
			Tuned Virtualization & 1000 & 1.219 & 2.62 & 3.988 & 0.812 \\ \hline
		\end{tabular}}
\end{table}

    \section{Discussion}
    \section{Summary and Outlook}
    Zusammenfassung stellt das Resümee und die kritische Reflexion über das Projekt dar.

    \section{Literaturverzeichnis}
    {
      \renewcommand{\refname}{\vspace{-5mm}} % keine eigene Überschrift durch \bibliography...
      \linespread{.8}\selectfont
      \bibliography{Literature_paper}
    }

>>>>

\usepackage{url} %Darstellung von URLs erlauben

%Deutsch (wahlweise mit DE)
\newcommand{\acessedthrough}{Verfügbar unter:}%Für URL-Angabe
\newcommand{\acessedthroughp}{Verfügbar durch:}%Für URL-Angabe (Geschützte Datenbank, Zugriff durch FH)
\newcommand{\acessedat}{Zugang am}%Für URL-Datum-Angabe
%Englisch (wahlweise mit EN)
%\newcommand{\acessedthrough}{Available at:}%Für URL-Angabe
%\newcommand{\acessedthroughp}{Available through:}%Für URL-Angabe (Geschützte Datenbank, Zugriff durch FH)
%\newcommand{\acessedat}{Accessed}%Für URL-Datum-Angabe
% bis hierher

@@

% own 
\usepackage{multirow}
\usepackage{array}
\usepackage[backend=biber, style=numeric, sorting=none]{biblatex}
\addbibresource{literature/Paper.bib}
% Set the bibliography font size to scriptsize
\AtBeginBibliography{\scriptsize}
\renewbibmacro{in:}{}

@@

\begin{multicols}{2}
  \abstract{\textbf{This thesis investigates the virtualization of a real-time operating system, which is built with Yocto, employs hard real-time through Xenomai 3 and is emulated using QEMU/KVM. The primary objective aim was to bridge the latency gap between virtualized and bare metal versions to ensure deterministic and reliable behavior, which is crucial for real-time robotics. Initial latency measurements showed a high gap between bare metal and virtualized setups. Extensive tuning across BIOS, kernel, host OS, QEMU/KVM, and guest OS reduced the worst-case latency from 707.622µs to 17.134µs, close to the bare metal performance of 10.709µs. The improvement was validated using a robotic application, comparing tuned virtualization with untuned and hardware versions.}} {\textbf{Virtualization, Real-Time Systems, Latency Reduction, Robot Control}}

  \section{Introduction}
  In today's industrial production and automation, robots must react to their environment and perform time-critical tasks within strict time constraints. Delays or errors can have catastrophic consequences in some cases. Virtualizing real-time operating systems for robotic control offers many advantages over traditional hardware-based approaches, especially when it comes to scaling~\cite{abbasiExploringOpenStackScalable2023}, flexibility and costs~\cite{taccariEmbeddedRealTimeVirtualization2014}. However, virtualization also introduces overhead and latency~\cite{casiniLatencyAnalysisVirtualization2021}~\cite{zhangEvaluatingOptimizingVirtualization2010}. In this master's thesis, the goal was the virtualization of a real-time operating system to control a robot, with a focus on compliance with real-time determinism. Specifically, the aim was to bring the latency of the virtualized Salamander 4 closer to that of the bare metal version, as initial latency measurements revealed a significant latency gap between the bare metal and virtualized setup.

  \section{Methodology}
  In order To address this gap, an extensive tuning process was carried out to achieve real-time performance and determinism. These tunings were added sequentially and tested together. This involves configurations spanning the BIOS, kernel, host operating system (OS) and QEMU/KVM virtualization layer, as demonstrated in~\cite{RealTimePerformanceTuning2022}. The guest OS is already based on Xenomai and therefore, no additional modifications were necessary. Ubuntu 22.04.4 LTS served as the host operating system, while Salamander 4, built with Yocto and based on Linux 5.15.94, was the guest OS. Xenomai provided real-time capabilities for the guest and its \texttt{latency} tool was used to measure the scheduling latency of the real-time OS. QEMU with KVM was employed for virtualization. Trace-cmd and Kernelshark were used for kernel tracing and visualization of trace data. The testing robot was connected to the virtual CPUS via a VARAN bus interface.

  \section{Implementation}
  The initial latency values were first measured with the \texttt{latency} program of the Xenomai tool suite. The tests were conducted with a sampling period of 100 µs, using a periodic user-mode task, and was assigned a priority of 99. If any sample's latency value was greater than 100µs, this was considered an overrun. There was a significant initial gap in latency statistics between the virtualized Salamander 4 and the Salamander 4 on bare metal. To address this gap, an extensive tuning process was carried out to achieve real-time performance and determinism. These modifications were added sequentially and tested together. Previous tunings were not reverted when moving to the next tunings. First, the BIOS was configured, followed by the kernel. The BIOS settings were not reverted when moving to the kernel configurations. Next, the host was configured, but the BIOS and kernel settings remained unchanged. The guest configurations were not changed since Xenomai already provides real-time capabilities. Finally, QEMU/KVM settings were applied. On top of that, a robotic application further confirmed the improvements. The difference between the command time and the time the signal reaches the PWM was measured 1,000 times for the untuned, tuned and hardware version of Salamander 4.

  \section{Results}
  This section summarizes the results after the real-time performance tunings, both in terms of the \texttt{latency} program and the robotic application. Figure~\ref{fig:max_latency_combined_results} displays the change in latency after each real-time performance tuning. Table~\ref{tab:latency_tables_combined} gives an overview of the most important metrics of the measurements after each tuning. 
\clearpage
  \begin{figure}[H]
    \centering
    \includegraphics[width=1.0\columnwidth]{img/results/gnuplot_combined_max_latency_all.png}
    \caption[Comparison of Latency Distribution of Salamander 4 Configurations]{Comparison of Latency Distribution of Salamander 4 under different Configurations}
    \label{fig:max_latency_combined_results}
  \end{figure}

  \begin{table}[H]
    \centering
    \footnotesize
    \caption{Comparison of Latency Statistics in Salamander 4 under different Configurations}
    \label{tab:latency_tables_combined}
    \begin{tabular}{|c|c|c|c|c|}
      \hline
      \multirow{2}{*}{\textbf{Tuning}} & \multicolumn{3}{c|}{\textbf{Latency (µs)}} & \multirow{2}{*}{\textbf{Overruns}} \\ \cline{2-4}
      & \textbf{Min} & \textbf{Avg} & \textbf{Max} & \\ \hline
      Bare Metal & 0.613 & 1.380 & 10.709 & 0 \\ \hline
      \begin{tabular}[c]{@{}c@{}}Untuned\\ Virtualization\end{tabular} & 2.536 & 8.940 & 707.622 & 43 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After BIOS\\ Configurations\end{tabular} & 0.969 & 3.948 & 457.545 & 94 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After Kernel\\ Configurations\end{tabular} & 2.545 & 4.811 & 21.694 & 0 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After Host\\ Configurations\end{tabular} & 2.591 & 4.834 & 18.441 & 0 \\ \hline
      \begin{tabular}[c]{@{}c@{}}After QEMU\\ Configurations\end{tabular} & 2.614 & 4.779 & 17.134 & 0 \\ \hline
    \end{tabular}
\end{table}

The improved latency was also tested with a robotic application. The difference between the command issuance time and the time the signal reaches the PWM was measured 1,000 times for the bare metal, untuned and tuned version of Salamander 4. The results are visualized in Table~\ref{tab:robotic_application_latency_values_combined}.

  \begin{table}[H]
    \centering
    \footnotesize
    \caption{Comparison of Latency Statistics in Salamander 4 under different Configurations}
    \label{tab:robotic_application_latency_values_combined}
    \begin{tabular}{|c|c|c|c|c|}
      \hline
      \multirow{2}{*}{\textbf{Tuning}} & \multicolumn{3}{c|}{\textbf{Latency (ms)}} & \multirow{2}{*}{\begin{tabular}[c]{@{}c@{}}\textbf{Std Dev}\\ \textbf{(ms)}\end{tabular}} \\ \cline{2-4}
      & \textbf{Min} & \textbf{Avg} & \textbf{Max} & \\ \hline
      Bare Metal & 1.211  & 1.347 & 1.49 & 0.082 \\ \hline
      \begin{tabular}[c]{@{}c@{}}Untuned\\ Virtualization\end{tabular} & 3.1 & 24.603 & 129.46 & 13.876\\ \hline
      \begin{tabular}[c]{@{}c@{}}Tuned\\ Virtualization\end{tabular} & 1.219  & 2.62 & 3.988 & 0.812 \\ \hline
    \end{tabular}
\end{table}

  \section{Discussion}
  The results shows that the real-time performance tunings applied to the virtualized Salamander 4 OS reduced latency and improved determinism. The maximum latency decreased from 707.622µs to 17.134µs after tuning BIOS, kernel, host, guest, and QEMU/KVM configurations. Especially the BIOS and kernel configurations played a crucial role in reducing the maximum latency, with a major fall to 21.694µs. This also eliminated overruns going forward. Host configurations, including
  the PREEMPT-RT patch, CPU and interrupt affinity, and real-time prioritization of QEMU further reduced latency a little bit down to 18.441µs. The guest OS already had real-time capabilities through Xenomai, hence the focus was primarily on optimizing the host and virtualization layer to achieve the desired real-time performance. Finally, QEMU/KVM configurations, such as tuning the LAPIC timer advance and using hugepages brought the latency down to the final worst latency value of 17.134µs. This value is very close to the bare metal value of 10.709\textmu s. The goal was set to achieving latency values below 50 microseconds in the duration of the measuremenet. This goal was acheived.
  A robotic application in a practical scenario validated the improvement of the tuned virtualization compared to the unmodified version. The difference between command time and signal reaching the PWM was measured 1,000 times for untuned, tuned, and hardware versions of Salamander 4. The worst latency dropped from 129ms to 3.988ms in the developed application. Compared to the hardware version's worst latency of 1.49ms, the tuned virtualization came very close to the determinism and reliability of the hardware.
  \section{Summary and Outlook}
  This master's thesis aimed to virtualize a real-time operating system for robot control, focusing on real-time determinism. The goal was to reduce latency in virtualized Salamander 4 to match its bare metal version. Initial tests showed a significant latency gap between virtualized and bare metal versions. Extensive tuning was performed sequentially: BIOS, kernel, host, and QEMU/KVM configurations. This process reduced worst latency from 707.62µs to 17.134µs, close to the bare metal value of 10.709µs, achieving the goal of sub-50µs latency.
  A robotic application confirmed these improvements. Measuring the time between command and PWM signal 1,000 times showed worst latency dropping from 129ms to 3.988ms, approaching the hardware version's 1.49ms.
  Future work could include further optimizations, testing other hypervisors, and evaluating performance under various workloads and stressed situations.
  Future work could include additional configurations and optimizations of the virtualization layer, investigating other hypervisors and virtualization technologies, and extensive testing under various workloads and stressed situations.

  \section{Bibliography}
  \printbibliography
